---
title: "DSCI401_HW2"
output: html_document
date: "2022-09-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

Using the Teams data frame in the Lahman package:

## part (a) (10 points) 

Create a data frame that is a subset of the Teams data frame that contains only the years from 2000 through 2009 and the variables yearID, W, and L.

```{r}
teams <- read.csv("Teams.csv")
sub_teams <- subset(teams[teams$yearID <= 2009 & teams$yearID >= 2000,], select = c("yearID", "W", "L"))
```

## part (b) (10points)

How many years did the Chicago Cubs (team ID is “CHN”) hit at least 200 HRs in a season and what was the median number of wins in those seasons.

```{r}
cubs <- teams[teams$teamID == "CHN",]

library(tidyverse)
cubs_runs <- cubs %>%
  group_by(yearID) %>%
  summarize(runs = sum(HR), wins = W)

high_runs <- cubs_runs[cubs_runs$runs >= 200,]
dim(high_runs)[1]
median(high_runs$wins)
```


## part (c) (10 points) 

Create a factor called election that divides the yearID into 4-year blocks that correspond to U.S. presidential terms. The first presidential term started in 1788. They each last 4 years and are still on the schedule set in 1788. During which term have the most home runs been hit?

```{r}
# create a new column for presidential term
# subtract year 1787, divide by 4, and round up
pres_terms = c()
for (y in teams$yearID) {
  pres_terms <- c(pres_terms, ceiling((y-1788)/4))
}

teams$pres_term <- pres_terms

# find total runs grouped by presidential term
most_runs <- teams %>%
  subset(select = c(pres_term, HR)) %>%
  group_by(pres_term) %>%
  summarize(runs = sum(HR))


# find the presidential term with max home runs
most_runs[most_runs$runs == max(most_runs$runs),]
```


## part (d) (10 points) 

Make a line plot of total home runs per season and stratify by league. Remove observations where league is missing.

```{r}
# drop lgID NA's
lg_teams <- 
  teams[!is.na(teams$lgID),] %>%
  group_by(lgID)

# create line plot
library(gcookbook)

ggplot(lg_teams, aes(x = yearID, y = HR, colour = lgID)) +
  geom_line()
```

## part (e) (10 points) 

Create an indicator variable called “winning record” which is defined as TRUE if the number of wins is greater than the number of losses and FALSE otherwise. Plot a scatter plot of Runs (R) vs Runs against (RA) with the color of each point showing whether that team had a winning record or not.

```{r}
library(dplyr)

# Adding winning record based on wins and losses
teams <- teams %>%
  mutate(winning_record = case_when(
    W > L ~ TRUE,
    W <= L ~ FALSE
    ))

# color selection
colors <- c("#FDAE61", # Orange
            "#D9EF8B") # Light green

# Scatter plot
plot(teams$RA, teams$R, pch = 19, col = colors[factor(teams$winning_record)], main = "Runs vs Runs Against by Winning Record", xlab = "Runs Against", ylab = "Runs")
```


# Question 2

Use the nycflights13 package and the flights data frame to answer the following questions:

## part (a) (10 points) 

What month had the highest proportion of cancelled flights? What month had the lowest? Interpret any seasonal patterns.

```{r}
# import the data
flights <- read.csv("https://raw.githubusercontent.com/gjm112/DSCI401/main/data/flights.csv")

# Adding cancelled column based on when air time is NA
flights <- flights %>%
  mutate(cancelled = case_when(
    is.na(air_time) == TRUE ~ 1,
    is.na(air_time) == FALSE ~ 0
    ))

# find proportion cancellations per month
month_cancels <- flights %>%
  subset(select = c(month, cancelled)) %>%
  group_by(month) %>%
  summarize(cancellations = mean(cancelled))

# find month with max proportion of cancellations
month_cancels[month_cancels$cancellations == max(month_cancels$cancellations),]

# find month with min proportion of cancellations
month_cancels[month_cancels$cancellations == min(month_cancels$cancellations),]
```

The Winter months, such as February, have the most cancellations, while October has the least amount of cancellations, perhaps because it is in the Fall


## part (b) (10 points) 

Given that a delay is longer than an hour, what is the average time of the total delay by airport (i.e. origin)

```{r}
# subset for delays longer than an hour
long_delays <- flights[flights$arr_delay > 60,]

# average delay by airport
avg_delays <- 
  long_delays %>%
  subset(select = c(origin, arr_delay)) %>%
  group_by(origin) %>%
  summarize(avg_delay = mean(arr_delay))
  
avg_delays
```


## part (c) (10 points) 

What is the average air time for all flights by carrier? Which carrier has the longest average air time on their flights?

```{r}
# drop air_time NA's
flights <- flights[!is.na(flights$air_time),]

# average air time by carrier
air_times <- flights %>%
  subset(select = c(carrier, air_time)) %>%
  group_by(carrier) %>%
  summarize(avg_air_time = mean(air_time))

air_times


# carrier with longest average air time
air_times[air_times$avg_air_time == max(air_times$avg_air_time),]
```


## part (d) (10 points) 

Keeping only flights that had a delay greater than 0, create a histogram for each month of the delay data.

```{r}
# subset for delays greater than 0
delays <- flights[flights$arr_delay > 0,]

# create histogram by month
library(ggplot2)
ggplot(delays, aes(x=arr_delay)) +
geom_histogram(position="identity") + facet_wrap(~ month) + ggtitle("Delay Times Grouped by Month")
```


## part (e) (10 points) 

Create side-by-side boxplots of delay times for flights with delays 60 minutes or greater for the top

```{r}
flight_counts <- table(long_delays$carrier)

# create a dictionary with total flights for each airline
top_airlines = c()
for (c in unique(long_delays$carrier)) {
  top_airlines[c] = flight_counts[c]
}

# sort the dictionary and extract top five airlines
top_carriers <- sort(top_airlines, decreasing = TRUE)[1:5]


# take subset of only top five airlines
airlines <- long_delays[long_delays$carrier %in% names(top_carriers), ]

# create boxplot grouped by airline
boxplot(airlines$arr_delay ~ airlines$carrier,
        col='steelblue',
        main='Delay Times by Airline',
        xlab='Airline',
        ylab='Delay Time') 
```

Google Colab Notebook: https://colab.research.google.com/drive/1wRpB2C5tIDuGMFqAZYS3gjPRVlx1wFYv?usp=sharing

